---
title: "GECO style map"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rgdal)  # for readOGR

## function for creating the color key
source("../R/plot_discrete_cbar.R")
```

## Example data

Obtain an example raster data as NetCDF file, here the file `cwdx80.nc`, from Zenodo:

*Stocker, Benjamin D. (2021). Global rooting zone water storage capacity and rooting depth estimates (v1.0) [Data set]. Zenodo. [https://doi.org/10.5281/zenodo.5515246](https://doi.org/10.5281/zenodo.5515246)*

Save it in `./data/`.

## Treating resolution

The file `cwdx80.nc` contains a map of the estimated root zone water storage capacity at 0.05˚ resolution. This is excessively high for a plot of a global map in a journal article (printed on less than A4 format). Therefore, we regrid the data to 0.5˚ first. We do this with the shell command `cdo remapbil` using [CDO](https://code.mpimet.mpg.de/projects/cdo/). The grid information file is provided here.
```{r eval=FALSE}
system("cdo remapbil,../data/gridfile_halfdeg.txt ../data/cwdx80.nc ../data/cwdx80_halfdeg.nc")
```

To facilitate things, the file `data/cwdx80_halfdeg.nc` is provided as part of this repository.

Read the lower resolution NetCDF file into an R data frame using the function `rbeni::nc_to_df()`.
```{r}
df <- rbeni::nc_to_df("../data/cwdx80_halfdeg.nc", varnam = "cwdx80") |> 
  mutate(lon = round(lon, digits = 2), lat = round(lat, digits = 2))
```

This is the starting point of our map plotting.

## Plot map

The workflow below shows how spatial data can be plotted. Here, spatial data is a data frame containing a column `lon` and a column `lat` for the longitude and latitude in degrees, respectively. the column `cwdx80` contains the values of the variable that is to be displayed and visualized by colors of pixels.

```{r}
## Reduce extent of map in latitude
latmin <- -60
latmax <- 80
lonmin <- -180 
lonmax <- 180

## Define bins of color scale
breaks <- c(seq(0, 100, by = 20), 150, 200, 300, 500, 700, 900, 1200, Inf)

## define domain object
domain <- c(lonmin, lonmax, latmin, latmax)

## read 110 m resolution coastline from NaturalEarth data (is a shapefile)
layer_coast <- readOGR(paste0("../data/ne_110m_coastline/ne_110m_coastline.shp"), 
                       paste0("ne_110m_coastline"))

## resize data frame to limit data volume (geographic clipping is done separately)
df <- df |> 
  dplyr::filter(lon > domain[1] & lon < domain[2] & lat > domain[3] & lat < domain[4])

## crop coastline data to domain
mycrop <- function(x, domain){

  # domain should be a vector of four values: c(xmin, xmax, ymin, ymax)
  x@data$id <- rownames(x@data)

  fortify(x, region="id") |> 
    as_tibble() |> 
    dplyr::left_join(x@data, by = "id") |> 
    dplyr::filter(long > domain[1] & long < domain[2] &
                    lat > domain[3] & lat < domain[4])
}
coast_crop <- mycrop(layer_coast, domain)
    
nbin <- length(breaks) - 1
breaks_with <- breaks

## Indicate that color scale extends to minus or plus infinity by a triangle
## instead of the top or bottom box of the color key.
toptriangle <- FALSE
bottomtriangle <- FALSE
if (is.infinite(breaks[length(breaks)])){
  toptriangle <- TRUE
  breaks <- breaks[-(length(breaks)-1)]
}
if (is.infinite(breaks[1])){
  bottomtriangle <- TRUE
  breaks <- breaks[-2]
}
nbin <- length(breaks) - 1

## add dummy rows to make sure values in layer span the entire range
df <- df |> 
  bind_rows(
    tibble(
      lon = NA,
      lat = NA,
      cwdx80 = breaks[1:(length(breaks)-1)] + 0.5 * (breaks[2]-breaks[1])
    )
  )

## make the data discrete
df$layercut <- as.factor(base::cut(df$cwdx80, breaks = breaks, labels = FALSE, include.lowest = TRUE))

## Define colors
colorscale <- viridis::cividis(nbin, direction = 1)

## or alternatively
# colorscale <- scico::scico(nbin, palette = "batlowK", direction = 1)

## Defiine color of the top and bottom triangle if required
if (toptriangle){
  colorscale <- c(colorscale, colorscale[length(colorscale)])
}
if (bottomtriangle){
  colorscale <- c(colorscale[1], colorscale)
}

## define the map theme 
## Not used below!
theme_map <- theme_grey() +    # theme_minimal()
  theme(

    plot.title = element_text(hjust = 0, face = "bold", size = 18),

    legend.position = "right", # c(0.07, 0.35), #"left"
    # legend.key.size = unit(c(5, 1), "mm"),
    legend.title=element_text(size=12),
    legend.text=element_text(size=10),

    # axis.line = element_blank(),
    # axis.text = element_blank(),
    # axis.title = element_blank(),

    # panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
    # plot.margin = unit( c(0, 0, 0, 5) , "mm")
  )

## create axis tick labels
lat_breaks <- seq(-90, 90, by = 30)
lon_breaks <- seq(-180, 180, by = 60)

lat_labels <- sapply( 
  lat_breaks, 
  function(x) if (x>0) {parse(text = paste0(x, "*degree ~ N"))} 
              else if (x==0) {parse(text = paste0(x, "*degree"))} 
              else {parse(text = paste0(-x, "*degree ~ S"))} 
  )
lon_labels <- sapply( 
  lon_breaks, 
  function(x) if (x>0) {parse(text = paste0(x, "*degree ~ E"))} 
              else if (x==0) {parse(text = paste0(x, "*degree"))} 
              else {parse(text = paste0(-x, "*degree ~ W"))} 
  )

## Create ggplot object
ggmap <- ggplot() +

  ## main raster layer
  geom_tile(data = df, aes(x = lon, y = lat, fill = layercut, color = layercut), show.legend = FALSE) +

  scale_fill_manual(values = colorscale) +
  scale_color_manual(values = colorscale) +
  xlab('') + 
  ylab('') +
  coord_sf(expand = FALSE) +
  theme_bw() +
  theme(axis.ticks.y.right = element_line(),
        axis.ticks.x.top = element_line(),
        panel.grid = element_blank()) +
  
  ## add coastline
	geom_path(data = coast_crop, aes(x = long, y = lat, group = group), color = 'gray25', size = 0.1) +
  
  ## add a title to the map
  labs(title = expression(paste(italic("S")[CWDX80])))

## use the function obtained from https://github.com/adrfantini/plot_discrete_cbar
## here, it's in the directory ./R/
gglegend <- plot_discrete_cbar(
  breaks           = breaks_with, # Vector of breaks. If +-Inf are used, triangles will be added to the sides   he color bar
  colors           = colorscale,
  legend_title     = "(mm)",
  legend_direction = "vertical",
  spacing = "constant",
  expand_size_y = 0.4,
  width = 0.03,
  font_size = 3  # of color key labels
  )

## Combine map and legend with cowplot
cowplot::plot_grid(ggmap, gglegend, ncol = 2, rel_widths = c(1, 0.15))
```

**Issues**

- longitude and latitude tick labels don't show the degrees sign
- would be nice to clip by coasline and give the ocean a different color
- demo in a separate section how to change projection to Robin
